---
title: "FjordLight Example"
author: "Bernard Gentili & Robert Schlegel"
date: "2023-10-19"
description: "This vignette walks the user through a viable workflow for using the functions within this package to work with the PAR dataset."
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{FjordLight Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
```

# Overview

The __`FjordLight`__ package has been built explicitly to facilitate the use of Arctic fjord PAR dataset developed via the [FACE-IT project](https://www.face-it-project.eu/). FACE-IT has received funding from the European Union's Horizon 2020 research and innovation programme under grant agreement no. 869154. This dataset was a development of that created by @gattuso2020global.

Below one will find a range of examples that should help to clarify the useful order of these functions in a workflow. All of these examples are also available within the help files for the given functions.

```{r setup}
library(FjordLight)
```

# Caveat emptor

With 64 GB of RAM (a typical desktop) your computer should have no problem loading the data, even with the most extensive fjords. With 4 GB (a typical laptop) you can only handle the smaller fjords: 'kong', 'young', and 'nuup'. However if you do not need the monthly `PARbottom` time series, the package should work with all fjords. In this case, when calling the `fl_LoadFjord()` function, set the `TS` argument to `FALSE` (this is the default value).

# Workflow

## List of available fjords
```{r eval=FALSE}
fl_ListFjords()
```

## Download a fjord

```{r eval=FALSE}
# Choose a fjord from the list above
fjord <- "kong"

# If the file has already been downloaded a message will be shown
fl_DownloadFjord(fjord)
```
To save the file to your computer (highly recommended):

```{r eval=FALSE}
# Note: this will require that a folder named 'data' exists in your current workig directory
# One may see one's working directory written in small text next to the version of R in the console pane
fl_DownloadFjord(fjord, dirdata = "data")
```

For the purposes of this vignette we will use the small toy dataset that is included within __`FJordLight`__:

```{r}
# Chose to load all of the monthly bottom PAR values or not
WANT_TIME_SERIES <- TRUE

# Load he data
fjord <- "test"
fjorddata <- fl_LoadFjord(fjord, dirdata = system.file("extdata", package = "FjordLight"), TS = WANT_TIME_SERIES)
str(fjorddata, list.len = 15)
```


## Get geographic parameters

```{r}
flget_geoparameters(fjorddata)
```

## Get bathymetry

```{r fig.width = 6, fig.height=6}
# All depths (what = "o" ; o for Ocean), as raster
all_bathy <- flget_bathymetry(fjorddata, what = "o", mode = "raster", PLOT = TRUE)

# Coastal zone [0-200m] (what = "c" ; c for Coastal), as raster
coastal_bathy <- flget_bathymetry(fjorddata, what = "c", mode = "raster", PLOT = TRUE)

# Shallow zone [0-50m] (what = "sl" ; s for Shallow, l to add land), as raster
shallow_land <- flget_bathymetry(fjorddata, what = "sl", mode = "raster", PLOT = TRUE)

# Just land; note the difference in colour palette
just_land <- flget_bathymetry(fjorddata, what = "l", mode = "raster", PLOT = TRUE)

# As a data.frame (mode = "df" : longitude, latitude, depth)
sea <- flget_bathymetry(fjorddata, what = "s", mode = "df", PLOT = FALSE)
cz <- flget_bathymetry(fjorddata, what = "c", mode = "df", PLOT = FALSE)

# you may add letter "l" if you want land elevation
sealand <- flget_bathymetry(fjorddata, what = "sl", mode = "df")

# Example of structure
str(sealand)
```

## Get climatologies of PAR0m, Kpar, and PARbottom

```{r fig.width = 6, fig.height=6}
# PAR0m and PARbottom for July
P07 <- flget_climatology(fjorddata, optics = "PAR0m", period = "Clim", month = 7, PLOT = TRUE)
print(P07)
Pb7 <- flget_climatology(fjorddata, optics = "PARbottom", period = "Clim", month = 7, PLOT = TRUE)
print(Pb7)

# PARbottom Global
PbG <- flget_climatology(fjorddata, optics = "PARbottom", period = "Global", PLOT = TRUE)
print(PbG)

# PAR0m, Kpar, and PARbottom for year 2012 as 3 columns data.frame
P02012 <- flget_climatology(fjorddata, optics = "PAR0m", period = "Yearly", year = 2012, mode = "df")
k2012 <- flget_climatology(fjorddata, optics = "Kpar", period = "Yearly", year = 2012, mode = "df")
Pb2012 <- flget_climatology(fjorddata, optics = "PARbottom", period = "Yearly", year = 2012, mode = "df")
head(Pb2012)
```

If you want the data as a data.frame:

```{r}
# first get pixels area
area <- flget_area(fjorddata, mode = "df")

# Then bind the data frames and remove rows with missing values
PAR_area <- cbind(sea, area[3], P02012[3], k2012[3], Pb2012[3])
PAR_area <- PAR_area[complete.cases(PAR_area),]
head(PAR_area)
```

## Get monthly time series of PARbottom

```{r fig.width = 6, fig.height=6}
# Years 2003 to 2004 - months July to August
mts <- flget_PARbottomMonthlyTS(fjorddata, month = 7:8, year = 2003:2004, PLOT = TRUE)
print(mts)

# Or as a data.frame
mts_2003 <- flget_PARbottomMonthlyTS(fjorddata, year = 2003, PLOT = FALSE, mode = "df")
head(mts_2003)
```

```{r eval=FALSE}
# All months - all years - as data.frame: 
  # columns = months (8, March to October) * years (20, 2003 to 2022) + 2 (lon lat) = 162
# NB: This may be too large for some laptops, proceed with caution
mts_full <- flget_PARbottomMonthlyTS(fjorddata, mode = "df", PLOT = FALSE)
```

## P-functions

```{r fig.width = 6, fig.height=6}
# One may create their own functions
fG <- flget_Pfunction(fjorddata, type = "coastal", period = "Global", plot = FALSE)
# Then you can use it; for instance :
irradiance_levels <- c(0.1, 1, 10)
fG(irradiance_levels)

# Or load the pre-calculated values as a 2 column data.frame
f2012 <- flget_Pfunction(fjorddata, type = "coastal", period = "Yearly", year = 2012, mode = "df")
str(f2012)

# Plot P-functions
fGlob <- flget_Pfunction(fjorddata, type = "coastal", period = "Global", PLOT = TRUE, lty = 1, col = 1, lwd = 2, 
                         Main = paste(fjord, "coastal P-functions"), ylim = c(0, 50))
```

# References
